%!TEX root = Proposal-PhD.tex
%
\chapter{Conclusion}%
\label{chap:conclusion}

\textsc{The complexity of control systems} at and beyond the state of the art is
producing proportionally complex software. Management of software complexity is
an ongoing issue within the software engineering industry. New software
development techniques and architectural styles are being practiced to improve
software quality as a means of controlling complexity.

The preceding document has outlined a course of research expanding modern
control techniques \emph{and} software designs that enable more rapid progress
is \ac{nmpc} control systems research. I desire to take lessons from the leading
edge of the software industry and from computer science to discover techniques
and best practices for designing software for \ac{nmpc} that is scalable,
distributable and parallelisable.

For the proposal, I outline three main areas of contribution:
\begin{enumerate}[label=(C\arabic*)]
  %
  \item the \acl{sdd}.
  %
  \item the software capital, and
  %
  \item novel theoretical research in the field of optimal control
  systems.
  %
\end{enumerate}

In a test-driven development process, I wish to develop a modern framework using
functional, reactive programming techniques and the modern \ac{cpp17} language.
The product of that endeavour covers the first two contributions.

The proposed framework is to be the basis for several research projects,
\textit{in potentiâ}. For example, one such project investigates the use of
artificial neural networks to accelerate the optimisation stage of the
algorithm, based on a sampling of the potential field which simulates obstacles
in state space. That project, and the others hold novelty in the field of
control systems research and represent progress in \ref{contrib:theory}. They
additionally demonstrate the usefulness of the framework which underlies them.
